---
title: "Capstone 2 analysis"
author: "Eugenio Mattei UMASS medical School"
date: "Run on `r Sys.Date()`"
documentclass: article
output: 
  fontsize: 14pt
  html_document:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval=TRUE, echo=FALSE, error=FALSE, warning = FALSE) 

rm(list=ls())

library(data.table)
library(ggplot2)
library(dplyr)
library(synapseClient)
library(edgeR)
library(preprocessCore)
library(wesanderson)

synapseLogin()

#counts syn11305250
#metadata syn11305353
Neurons="#0072B2"
NonNeurons="#F0E442"
CNON="#FF7903"
ACC="#009E73"
PFC="#CC79A7"
ON="#AEAFAE"
#brainregions.col=c(ACC,PFC,ON)
#celltype.col=c(Neurons,NonNeurons,CNON)
celltype.col = c("NeuN+"=Neurons,"NeuN-"=NonNeurons,"CNON"=CNON)
brainregions.col=c("ACC"= ACC,"DLPFF" = PFC,"ON" = ON)

brainregions=c("Anterior\nCingular\nCortex", "Cerebellum", "Dorsolateral\nPre-Frontal\nCortex","Olfactory\nNeuroepithelium","Pre-Frontal\nCortex","Temporal\nCortex")
marks.col=c("#00B0F0","#FFCD00","#FF0000")

tol21rainbow= c("#771155", "#AA4488", "#CC99BB", "#114477", "#4477AA", "#77AADD", "#117777", "#44AAAA", "#77CCCC", "#117744", "#44AA77", "#88CCAA", "#777711", "#AAAA44", "#DDDD77", "#774411", "#AA7744", "#DDAA77", "#771122", "#AA4455", "#DD7788")
rich12equal = c("#000040", "#000093", "#0020E9", "#0076FF", "#00B8C2", "#04E466", "#49FB25", "#E7FD09", "#FEEA02", "#FFC200", "#FF8500", "#FF3300")
mycols=c("dodgerblue2", "#E31A1C", "green4", "#6A3D9A", "#FF7F00", "black",
         "gold1", "skyblue2", "#FB9A99", "palegreen2", "#CAB2D6", "#FDBF6F",
         "gray70", "khaki2", "maroon", "orchid1", "deeppink1", "blue1",
         "steelblue4", "darkturquoise", "green1", "yellow4", "yellow3",
         "darkorange4", "brown")

theme.common.long.label = theme(axis.text.x = element_text(face="bold", 
                                   size=14,
                                   angle=45,
                                   hjust=1,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.text.y = element_text(face="bold",
                                   size=14,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.title.y = element_text(face="bold",
                                    size=14),
        legend.text = element_text(face="bold",size=12),
        legend.title = element_blank(),
        axis.ticks.length=unit(-0.15, "cm"))

theme.common.short.label = theme(axis.text.x = element_text(face="bold", 
                                   size=14,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.text.y = element_text(face="bold",
                                   size=14,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.title.y = element_text(face="bold",
                                    size=14),
        legend.text = element_text(face="bold",size=12),
        legend.title = element_blank(),
        axis.ticks.length=unit(-0.15, "cm"))

#png("file.png",width=1920,height=1080,res=150)

```


Loading the Capstone2 metadata
```{r}
capstone2.metadata = fread(getFileLocation(synGet( 'syn11305353' )), header=TRUE, stringsAsFactors=FALSE,sep="\t",data.table = F) 
capstone2.metadata$has_GENOTYPE_bool=TRUE
capstone2.metadata$has_GENOTYPE_bool[capstone2.metadata$has_GENOTYPE=="FALSE"]=FALSE
head(capstone2.metadata)
```

*ChIPseq Statistics*

Overview of the number of sampler per __Study__
```{r}

temp = capstone2.metadata %>% group_by(Study,Mark) %>% summarise(n=n())

ggplot(temp,aes(as.factor(Study),n,fill=Mark))+
  geom_col(position='dodge')+
  theme_bw()+
  scale_fill_manual(values = marks.col,legend)+
  ggtitle("Number of Samples Per Study")+
  geom_text(data=temp, aes(y=n,label = n), size = 5,position = position_dodge(width = 0.90),vjust=-0.4,color="black")+
  labs(y = "Number of samples",x="")+
  theme(axis.text.x = element_text(face="bold", 
                                   size=14,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.text.y = element_text(face="bold",
                                   size=14,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.title.y = element_text(face="bold",
                                    size=14),
        legend.text = element_text(face="bold",size=12),
        legend.title = element_blank(),
        axis.ticks.length=unit(-0.15, "cm"))

```

Overview of the number of sampler per __Brain Region__
```{r}
temp = capstone2.metadata %>% group_by(Brain_Region,Mark) %>% summarise(n=n())

ggplot(temp,aes(as.factor(Brain_Region),n,fill=Mark))+
  geom_col(position='dodge')+
  theme_bw()+
  scale_fill_manual(values = marks.col)+
  ggtitle("Number of Samples Per Brain Region")+
  scale_x_discrete(name = "", labels=brainregions)+
  labs(y = "Number of samples")+
  geom_text(data=temp, aes(y=n,label = n),
            size = 5,
            position = position_dodge(width = 0.90),
            vjust=-0.3,
            colour="black")+
  theme(axis.text.x = element_text(face="bold", 
                                   size=14,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.text.y = element_text(face="bold",
                                   size=14,
                                   margin = unit(c(0.5,0.5,0.5,0.5), "cm")),
        axis.title.y = element_text(face="bold",
                                    size=14),
        legend.text = element_text(face="bold",size=12),
        legend.title = element_blank(),
        axis.ticks.length=unit(-0.15, "cm"))

```

Overview of the number of sampler per __Brain Region__ and __Cell Type__
```{r}
temp = capstone2.metadata %>% group_by(Brain_Region,Cell_Type,Mark) %>% dplyr::summarise(n=n())

ggplot(temp,aes(as.factor(Cell_Type),n,fill=Mark))+
  geom_col(position='dodge')+
  theme_bw()+
  scale_fill_manual(values = marks.col)+
  ylab("Number of samples")+
  xlab("")+
  ggtitle("Peaks Summary")+
  geom_text(data=temp, aes(y=n,label = n), size = 5,position = position_dodge(width = 0.90),vjust=-0.4,color="black")+
  facet_wrap(~Brain_Region)+ theme.common.long.label

```

Number of Sampled per __Cell Type__
```{r}
temp = capstone2.metadata %>% group_by(Cell_Type,Mark) %>% summarise(n=n())

ggplot(temp,aes(as.factor(Cell_Type),n,fill=Mark))+
  geom_col(position='dodge')+
  theme_bw()+
  scale_fill_manual(values = marks.col)+
  ylab("Number of samples")+
  xlab("")+
  ggtitle("Peaks Per Cell Type")+
  geom_text(data=temp, aes(y=n,label = n), size = 5,position = position_dodge(width = 0.90),vjust=-0.4,color="black")


temp = capstone2.metadata %>% group_by(Brain_Region,Diagnosis) %>% summarise(n=n())

ggplot(temp,aes(as.factor(Brain_Region),n,fill=Diagnosis))+
  geom_col(position='dodge')+
  theme_bw()+
  scale_fill_manual(values = wes_palette(n=4, name="GrandBudapest"))+
  scale_x_discrete(name = "", labels=brainregions)+
  labs(y = "Number of samples")+
  ggtitle("Number of Samples Per Diagnosis")+
  geom_text(data=temp, aes(y=n,label = n), size = 5,position = position_dodge(width = 0.90),vjust=-0.4,color="black")+
  theme.common.short.label

```

*ATACseq Summary*
```{r}
temp = capstone2.metadata %>% group_by(Brain_Region,has_ATACseq) %>% summarise(n=n())

ggplot(temp,aes(as.factor(Brain_Region),n,fill=has_ATACseq))+
  geom_col(color=c("black"))+
  theme_bw()+
  scale_fill_manual(values = c("white","darkgreen"),name="Has ATACseq")+
  scale_x_discrete(name = "", labels=brainregions)+
  labs(y = "Number of samples")+
  ggtitle("ATACseq Statistics")+
  geom_label(data=temp, aes(y=n,colour=has_ATACseq,label = n), size = 5,vjust=-0.5,color="black",show.legend = FALSE)+
  theme.common.short.label

```

*RNAseq summary*
```{r}
temp = capstone2.metadata %>% group_by(Brain_Region,has_RNAseq) %>% summarise(n=n())

ggplot(temp,aes(as.factor(Brain_Region),n,fill=has_RNAseq))+
  geom_col(col="black")+
  theme_bw()+
  scale_fill_manual(values = c("white","darkgreen"),name="Has RNAseq")+
  scale_x_discrete(name = "", labels=brainregions)+
  labs(y = "Number of samples")+
  ggtitle("RNAseq Statistics")+
  geom_label(data=temp, aes(colour=has_RNAseq,label = n), size = 5,vjust=-0.1,position = "stack",color="black",show.legend = FALSE)+
  theme.common.short.label

```

*Genotype summary*
```{r}
temp = capstone2.metadata %>% group_by(Brain_Region,has_GENOTYPE_bool) %>% summarise(n=n())

ggplot(temp,aes(as.factor(Brain_Region),n,fill=has_GENOTYPE_bool))+
  geom_col(col="black")+
  theme_bw()+
  scale_fill_manual(values = c("white","darkgreen"),name="Has Genotype")+
  scale_x_discrete(name = "", labels=brainregions)+
  labs(y = "Number of samples")+
  ggtitle("Genotype Statistics")+
  geom_label(data=temp, aes(colour=has_GENOTYPE_bool,label = n), size = 5,position = "stack",vjust=-0.1,color="black",show.legend = FALSE)+
  theme.common.short.label

```



#Counts Stats
```{r}
capstone2.h3k4me3.counts = fread(getFileLocation(synGet( 'syn11305250' )), header=TRUE, stringsAsFactors=FALSE,sep="\t",data.table = F) 
capstone2.h3k4me3.peaks  = fread(getFileLocation(synGet( 'syn11305254' )), header=TRUE, stringsAsFactors=FALSE,sep="\t",data.table = F) 
capstone2.h3k4me3.peaks$widths = capstone2.h3k4me3.peaks$End - capstone2.h3k4me3.peaks$Start

peak.stats= fread("~/Dropbox/psychENCODE-Eu/capstone2/H3K4me3/stats/peaks_stats_H3K4me3",h=F,stringsAsFactors = F)
dim(peak.stats)
head(peak.stats)
boxplot(peak.stats$V3)

rownames(capstone2.h3k4me3.counts)=capstone2.h3k4me3.counts$Geneid
capstone2.h3k4me3.counts=capstone2.h3k4me3.counts[,-1]


#4 samples missing
counts.reduced = capstone2.h3k4me3.counts[,names(capstone2.h3k4me3.counts) %in% capstone2.metadata$Sample_ID]
capstone2.metadata.sorted=capstone2.metadata[match(names(counts.reduced),capstone2.metadata$Sample_ID),]
head(capstone2.metadata.sorted)
studies = unique(capstone2.metadata.sorted$Study)

epimap.filter = rowSums(cpm(counts.reduced[,colnames(counts.reduced) %in% capstone2.metadata.sorted$Sample_ID[capstone2.metadata.sorted$Study==studies[1]]])>1) >= length(capstone2.metadata.sorted$Individual_ID[capstone2.metadata.sorted$Study==studies[1]])*0.10

epidiff.filter = rowSums(cpm(counts.reduced[,colnames(counts.reduced) %in% capstone2.metadata.sorted$Sample_ID[capstone2.metadata.sorted$Study==studies[2]]])>1) >= length(capstone2.metadata.sorted$Individual_ID[capstone2.metadata.sorted$Study==studies[2]])*0.10

cnon.filter = rowSums(cpm(counts.reduced[,colnames(counts.reduced) %in% capstone2.metadata.sorted$Sample_ID[capstone2.metadata.sorted$Study==studies[3]]])>1) >= length(capstone2.metadata.sorted$Individual_ID[capstone2.metadata.sorted$Study==studies[3]])*0.10

D = DGEList(counts = counts.reduced) ;
D = calcNormFactors(D);
#keep <- rowSums(cpm(counts.reduced)>1) >= 5;
keep = epimap.filter | epidiff.filter | cnon.filter
D = D[keep,];
D = calcNormFactors(D)
R = rpkm(D,gene.length = capstone2.h3k4me3.peaks$widths[capstone2.h3k4me3.peaks$Peak_name%in%rownames(D$counts)], log = TRUE)

```

##Clustering to test for separation between Cell Types

```{r,message=F,fig.width=9,fig.height=12}
library(devtools)
library(RColorBrewer)
library(rafalib)

#Load latest version of heatmap.3 function
hmcol <- colorRampPalette(brewer.pal(9, "GnBu"))(100)

cols <- brainregions.col[as.fumeric(as.character(capstone2.metadata.sorted$Brain_Region))]
cols2 <- celltype.col[as.fumeric(as.character(capstone2.metadata.sorted$Cell_Type))]
cols3 <- c("darkred","darkgreen","black")[as.fumeric(as.character(capstone2.metadata.sorted$Study))]
cols2

colLab=t(cbind(cols,cols2,cols3))
rownames(colLab)=c("brainRegion","cellType","Study")
#colnames(colLab)=names(capstone2.h3k4me3.counts)
colnames(colLab)=names(counts.reduced)
colLab
test_cormat <- round(cor(R,method = "spearman"),2)
test_dist = as.dist((1-test_cormat)/2)

test_col <- data.frame(row.names = colnames(R), 
                      Cell_Type = factor(capstone2.metadata.sorted$Cell_Type,levels=unique(capstone2.metadata.sorted$Cell_Type)))
test_colors <- list(Cell_Type = celltype.col[unique(test_col$Cell_Type)])
names(test_colors$Cell_Type) <- unique(test_col$Cell_Type)


groups = as.data.frame(cutree(hc,h=0.36))
names(groups)="clusters"

test_col$clusters=groups$clusters

pheatmap(
  mat               = test_dist,
  color             = hmcol,
  border_color      = NA,
  show_colnames     = FALSE,
  show_rownames     = FALSE,
  annotation_col    = test_col,
  annotation_colors = test_colors,
  drop_levels       = TRUE,
  fontsize          = 14,
  main              = "TEST"
)

hc = hclust(test_dist)
hc$order

hc.order = hc$labels[hc$order]
hc.df = data.frame(row.names = hc.order, len=peak.stats[match(hc.order,peak.stats$V2),V3])
hc.df$label=capstone2.metadata.sorted[match(hc.order,capstone2.metadata.sorted$Sample_ID),]$Cell_Type
hc.df
```
PCA

```{r}
cols <- brainregions.col[as.fumeric(as.character(capstone2.metadata.sorted$Brain_Region))]
cols2 <- celltype.col[as.fumeric(as.character(capstone2.metadata.sorted$Cell_Type))]
cols3 <- c("red","darkgreen","black")[as.fumeric(as.character(capstone2.metadata.sorted$Study))]
cols4 <- wes_palette(n=3,"GrandBudapest")[as.fumeric(as.character(capstone2.metadata.sorted$Diagnosis))]

counts.pca <- prcomp(t(R), center = TRUE, scale. = TRUE)


#png("Images/PCA_1_vs_2.png",width=1080,height=1080,res=90)
par(mfrow=c(2,2))
plot(counts.pca$x[,1:2], col = cols,pch=19,cex=1.5,ann = T,main="PCA 1 Vs. 2")
legend(x=-900,y=410, inset=.05, c("ACC","DLPFC","CNON"), fill=unique(cols), horiz=F,border = "white",box.col = "white")

plot(counts.pca$x[,1:2], col = cols2,pch=19,cex=1.5,ann = T,main="PCA 1 Vs. 2")
legend(x=-900,y=410, inset=.05, c("NeuN+","NeuN-","ON"), fill=unique(cols2), horiz=F,border = "white",box.col = "white")

plot(counts.pca$x[,1:2], col = cols3,pch=19,cex=1.5,ann = T,main="PCA 1 Vs. 2")
legend(x=-900,y=410, inset=.05, unique(capstone2.metadata.sorted$Study), fill=unique(cols3), horiz=F,border = "white",box.col = "white")

plot(counts.pca$x[,1:2], col = cols4,pch=19,cex=1.5,ann = T,main="PCA 1 Vs. 2")
legend(x=-900,y=410, c("Control","SCZ","Control:depression"), fill=unique(cols4), horiz=F,border = "white",box.col = "white")

#dev.off()
```

PCA 2 Vs 3

```{r}

#png("Images/PCA_2_vs_3.png",width=1080,height=1080,res=90)
par(mfrow=c(2,2))

plot(counts.pca$x[,2:3], col = cols,pch=19,cex=1.5,ann = T,main="PCA 2 Vs. 3")
legend("top", inset=.05, c("ACC","DLPFC","CNON"), fill=unique(cols), horiz=F,border = "white",box.col = "white",ncol= 3)

plot(counts.pca$x[,2:3], col = cols2,pch=19,cex=1.5,ann = T,main="PCA 2 Vs. 3")
legend("top", inset=.05, c("NeuN+","NeuN-","ON"), fill=unique(cols2), horiz=F,border = "white",box.col = "white",ncol= 3)

plot(counts.pca$x[,2:3], col = cols3,pch=19,cex=1.5,ann = T,main="PCA 2 Vs. 3")
legend("top", inset=.05, unique(capstone2.metadata.sorted$Study), fill=unique(cols3), horiz=F,border = "white",box.col = "white",ncol= 3)

plot(counts.pca$x[,2:3], col = cols4,pch=19,cex=1.5,ann = T,main="PCA 2 Vs. 3")
legend("top",inset=.05, c("Control","SCZ","Control:depression"), fill=unique(cols4), horiz=F,border = "white",box.col = "white",ncol= 3,text.width = 100)

#dev.off()

```


MDS
```{r}
fit <- cmdscale(test_dist, eig = TRUE, k = 2)

#png("Images/MDS.png",width=1080,height=1080,res=90)
par(mfrow=c(2,2))
plot(fit$points[,1:2], col = cols,pch=19,cex=1.5,ann = T,main="MDS")
legend("bottomleft", inset=.05, c("ACC","DLPFC","CNON"), fill=unique(cols), horiz=F,border = "white",box.col = "white")

plot(fit$points[,1:2], col = cols2,pch=19,cex=1.5,ann = T,main="MDS")
legend("bottomleft", inset=.05, c("NeuN+","NeuN-","ON"), fill=unique(cols2), horiz=F,border = "white",box.col = "white")

plot(fit$points[,1:2], col = cols3,pch=19,cex=1.5,ann = T,main="MDS")
legend("bottomleft", inset=.05, unique(capstone2.metadata.sorted$Study), fill=unique(cols3), horiz=F,border = "white",box.col = "white")

plot(fit$points[,1:2], col = cols4,pch=19,cex=1.5,ann = T,main="MDS")
legend("bottomleft",inset = .05, c("Control","SCZ","Control:depression"), fill=unique(cols4), horiz=F,border = "white",box.col = "white")
#dev.off()

```
